### 2024-04-30 Paper/paper.tex
Updates from Overleaf

    \begin{abstract}
		{\small
        Nonprofit firms have diverse goals and missions that make understanding their true objectives and behaviors difficult. A setting where this is particularly relevant is that of nonprofit hospitals, which make up the majority of hospitals in the US, yet there is mixed evidence on what drives their behavior. In some ways they seem to act like for-profits, while in other ways they may care more about prestige or quantity. In this paper, I investigate whether the leadership team composition of nonprofit hospitals is correlated with more for-profit behavior. I use a novel data set on nonprofit hospital executives in the US from 2009-2015 to investigate whether having clinical experience on a hospital executive team affects hospital behavior. In particular, I leverage several pay-for-performance initiatives in the US in 2012, and estimate the difference in response between for-profits, nonprofits, and clinical nonprofits. I find that while readmissions decreased on average for all types of hospitals, non-clinical executive team hospital decreased readmissions at a faster rate than those with a clinical executive team, and behaved more similarly to for-profit hospitals. The difference equates to approximately 15 additional readmissions for clinical team hospitals.

		} 
	\end{abstract}

@@

  The objectives and behaviors of firms that are not classic for-profits are not easily determined, and an extensive literature has sought to understand these types of firms. Nonprofits are thought to gain utility not solely from monetary profit, but also through accomplishing mission-based goals. Researchers have speculated a variety of objective functions that could motivate nonprofits such as maximizing prestige, income, or a quality/quantity trade-off (\cite{steinberg1986revealed}). An important contribution to this literature is the use of empirical strategies to analyze how firm behavior reveals attributes of the underlying objective function (\cite{sloan2000not}). In this paper, I leverage a large-scale shock to hospitals in the US to investigate whether differences in leadership composition reveal characteristics of hospital objective functions, namely the weight placed on profit vs. utility gained from benefiting society.

  Hospitals in particular have been a common focus when seeking to understand nonprofit firms. This is because, first, there are different hospital ownership types within the industry, giving researchers the ability to compare the behaviors of nonprofits with that of for-profits. Second, nonprofit hospitals make up 50\% of all hospitals in the US, and staff, on average, 207 beds. For comparison, for-profits make up around 36\% of hospitals and staff 107 beds (\cite{ASPE_2023}). Thus, nonprofit behavior in the hospital industry is relevant and directly affects consumers of health care. However, it remains unclear what type of objective function drives nonprofit hospitals (\cite{erus2002inferring}; \cite{deneffe2002not}; \cite{horwitz2009hospital}). In particular, the extent to which nonprofit hospitals gain utility from bettering society (e.g. increased charity care, high quality of care, etc.) is unknown.

  What has not been considered is how the leadership team within nonprofits potentially drives objectives. In large, publicly traded firms, executive characteristics are shown to be correlated with firm performance, indicating that there is more to firm behavior than for-profit or not-for-profit status (\cite{bertrand2003managing}; \cite{matsa2013female}; \cite{ahern2012changing}). Within nonprofit hospitals, there is variation in the propensity to hire executives with a clinical background, which could reasonably change how much weight the hospital places on profit vs. quality of care when making decisions. I contribute to our understanding of hospital and nonprofit behaviors by investigating whether clinical experience on an executive team affects hospital behavior after being financially penalized.

  Using publicly available tax forms, I construct a novel data set of nonprofit hospital executives from 2009-2015. I link hospitals in this data to characteristics in the American Hospital Association survey and Hospital Compare data. In order to capture the effect of clinical executives, not just their selection into certain types of hospitals, I consider the sample of hospitals whose leadership team remains stable over time, and characterize them as having clinical executive experience on their team or not, a time-invariant categorization. I then leverage the Hospital Readmissions Reduction Program (HRRP) of 2012. This program reduced Medicare payments to hospitals with above average readmission rates, and is a shock that is plausibly uncorrelated with previously determined leadership teams.

  I estimate the difference in readmission and mortality rates between hospitals with and without clinical experience before and after the penalties began. While readmissions directly affects the likelihood of being penalized, mortality rates are an explicit quality measure unrelated to the penalties. Including this as an outcome can help distinguish whether hospitals are improving quality or only taking measures to reduce readmissions. Not all hospitals were penalized, but all hospitals (with few exceptions) could have been penalized if readmission rates were high enough. Thus, I present estimates first for all hospitals. However, these estimates are biased if I am simply capturing the effect of clinical experience on already being a good quality hospital. Thus, I present two further specifications: one limiting to only hospitals who were eventually penalized, and another where I limit the sample further to hospitals which have an observably similar match.

  I assume that before the penalty, clinical and non-clinical team hospitals behaved similarly along quality dimensions, and that the composition of hospital executive teams determined prior to the penalties is exogenous. Under these assumptions, I identify the effect of having a clinical executive on hospital response to the penalty. I find that while both types of hospitals, those with clinical executives and those without clinical executives, decrease readmissions after being penalized, non-clinical team hospitals decrease readmissions at a faster rate. The difference equates to approximately 15 additional readmission cases per year for clinical teams, and there is no evidence that this result is driven by differences in readmissions prior to penalties. However, I find no differential effect of the program on mortality rates.

    This paper contributes to three strands of literature. First, it adds to our understanding of what underlies nonprofit hospital behavior. This literature is extensive, and a review of it is beyond the scope of this paper. To summarize, there are various aspects of nonprofit behavior and objectives for which the evidence of how much it differs from for-profits remains mixed: costs, uncompensated care, technology adoption, and quality (\cite{sloan2000not}; \cite{eggleston2008hospital}; \cite{moscelli2018effect}; \cite{moscone2020public}). What underlies this is the need to understand the ways in which nonprofits make choices differently than for-profits. To this point, researchers have found that nonprofits do not act as purely profit maximizers, but maximize some combination of profit and output (\cite{deneffe2002not}, \cite{chang2011nonprofit}). A paper with an approach similar to mine which investigates hospital behavior is \citeauthor{chang2011nonprofit} (\citeyear{chang2011nonprofit}), which looks at the effect of a cost shock on likely of shutting down and mix of profitable services. They find that nonprofits and for-profits are equally as likely to shut down after the shock, but only nonprofits adjust their mix of profitable services. Similarly, I investigate a financial shock to hospitals, but I look at the differential impacts of leadership teams within nonprofits. This paper adds to our understanding of inputs into nonprofit hospital objective functions, and whether this is dependent on leadership compositions.

    Second, many researchers have documented a correlation between executive backgrounds and firm performance specifically for CEOs, starting with \citeauthor{bertrand2003managing} (\citeyear{bertrand2003managing}), who show that manager fixed effects are an important driver of many firm behaviors and decisions. Female board members are associated with better oversight and more women executives (\cite{matsa2011chipping}; \cite{adams2009women}), but are negatively correlated with firm value (\cite{ahern2012changing}). Having more female executives is correlated with female employee wages and corporate strategies (\cite{flabbi2019female}; \cite{matsa2013female}); young male CEOs tend to be more aggressive in mergers and acquisitions while those with military experience are less aggressive (\cite{levi2010deal}; \cite{benmelech2015military}); CEOs with general ability tend to receive higher pay and perform better (\cite{kaplan2012ceo}; \cite{custodio2013generalists}, \cite{adams2018director}; \cite{frydman2019rising}). Further, Chief Diversity Officers are not found to have any effect on hiring more diversely in universities (\cite{bradley2022impact}). Due to data limitations, these studies focus mostly on large, publicly traded firms in the US. However, one study uses data from England to study whether CEOs affect hospital production, and find no association (\cite{janke2019impact}). I contribute to this literature first by investigating a context which is not largely studied but is policy relevant, nonprofit US hospitals, and considering a unique executive characteristic pertinent to that context. Second, I contribute by addressing leadership team endogeneity concerns through leveraging a shock to hospitals and strategically limiting the sample.

    Finally, I contribute to our understanding of how providers respond to pay-for-performance incentives in health care. The most in depth study of how hospitals respond to HRRP is \citeauthor{gupta2021impacts} (\citeyear{gupta2021impacts}). The author finds that hospitals decreased readmissions by 5\% and mortality rates by 2\% on average as a result of the program, confirming prior studies (\cite{mellor2017does}; \cite{ziedan2018essays}; \cite{ody2019decreases}; \cite{gupta2021impacts}). Importantly, around 40\% of the decrease in readmissions is due to selective patient practices. This raises important questions about how to incentivize improvement in quality of care without also incentivizing gamesmanship. This paper addresses a necessary first step in considering whether leadership compositions could affect likelihood of gaming, which is to first investigate whether leadership matters to hospital decisions at all.

@@

    The Hospital Readmissions Reduction Program (HRRP) was passed as a part of the Affordable Care Act in 2010 as a pay-for-performance incentive for hospitals to increase quality. In 2011, the Center for Medicare and Medicaid Services (CMS) released a set of rules that mandated penalties for hospitals with above average readmission rates. A readmission is when a patient returns to the hospital within 30 days of being discharged from a previous stay; avoidable readmissions are a bad outcome for patients and increase health care spending. The goal of HRRP is to lower readmissions through better care coordination, less initial stay complications, and better post-care instructions. Beginning in October 2012, hospitals with higher readmission rates than the national average in pneumonia, heart failure, or AMI (after adjusting for demographic characteristics) receive a lower reimbursement rate for Medicare patients. In 2015, CMS also included chronic obstructive pulmonary disease, coronary artery bypass graft surgery, and elective primary total hip arthroplasty and/or total knee arthroplasty as conditions which go into the penalty calculation.

    Penalized hospitals pay a flat rate for every Medicare patient, not just those for which they have too high readmissions. Further, CMS does not distinguish a necessary readmission vs. an avoidable readmission, any repeat hospital visit is included in the penalty calculation. Excess readmission rates are calculated using a rolling look-back period of 3 years to determine whether the hospital is penalized. Therefore, hospitals had incentive to react immediately once details of the program were announced in October of 2011. These penalties are not insignificant; penalized hospitals paid, on average, 4-5\% of revenue.

@@

    Nonprofits are governed by a board of directors, whose role is to set high level goals and strategies, and provide general oversight. The board is not in charge of day-to-day operations, but the general direction of the firm. They select members of the executive team to carry out day-to-day operations. While hospital executives typically do specialize by getting degree(s) in healthcare management or an MBA specific to healthcare, they do not often have hands on clinical experience. When there is a hospital executive with clinical experience, they likely decided on that career path after becoming a doctor. While some doctors earn additional degrees before stepping into an executive role, this is not a necessary condition to becoming a physician executive. An executive team usually consists of at least a CEO and CFO. Every nonprofit is different in exactly how they structure an executive team. Some nonprofits hire Chief Medical Officers, Chief Quality Officers, executive directors, vice presidents of various departments, etc. I consider all of these employees as part of the day-to-day management of the hospital and include them as a part of the executive team.

    All tax-exempt organizations in the US are required to file a Form 990 with the IRS each year. There are different types of forms, but any organization that grosses over \$200,000 must file the most extensive form 990. Sections of the form include a statement of revenue, statement of functional expenses, a balance sheet, and, as used in this project, a list of all key employees, executives, and board members. In this section, each firm is required to report the name and title, average hours per week, position, and compensation of their board members and executive leaders. Since these firms are tax exempt, their financial information is made publicly available through the public housing of these forms.

@@

    To characterize hospitals as having clinical experience on their executive team or not, I construct a novel data set of nonprofit hospitals in the US that contains names and select characteristics of executives tied to the hospital from 2009-2015. I gather this data from each hospital's publicly available Tax Form 990, which every sufficiently large nonprofit files each year in the US. To my knowledge, this is the first large scale gathering of leadership names from these forms.

    The historical tax forms for all nonprofits are housed by ProPublica\footnote{https://projects.propublica.org/nonprofits/}. I use the NonProfit Explorer API to extract the Employee Identification Numbers (EIN) of all nonprofits classified as a hospital according to their National Taxonomy of Exempt Entities (NTEE) code. After filtering out associations and specialty hospitals, there are around 3,000 firms left. Importantly, I need to be able to link hospitals in the tax forms to other hospital characteristics, including penalty status and readmission rates, in the American Hospital Association annual survey and Hospital Compare data. No official crosswalk between EINs and AHA IDs or Medicare numbers exists, so I must rely on matching by location and name of hospital in both data sets. After limiting to only general, nonprofit hospitals in the AHA survey, I confidently match approximately 1200 EINs to an AHA identification number based on exact name matches within the same state.

    For each of the 1200 EINs, I extract web URLs to Form 990 PDFs in each year. I download these locally and use optical character recognition (OCR) text extraction methods to scrape the relevant section of the pdf, which is titled ``Officers, Directors, Trustees, Key Employees, and Highest Compensated Employees". Finally, I use string cleaning methods to identify names and positions for each EIN, year. I also identify titles indicating a physician: md, dr., or do. Using OCR extraction can be unreliable in some instances when the text is handwritten or the font is too small. Therefore, I manually find names, positions, and titles for observations that are missing after the initial extraction. This yields 850 EINs that are matched to AHA data, and are not missing text information.

    Due to the lack of consistency in organization of executive teams, I focus mainly on the executive teams as a whole instead of specific positions. I drop anyone who is solely a board member, and focus on those who have executive in their title or are labeled as a president or vice president of a specific aspect of the hospital. My identification strategy relies on hospitals that have stable executive leadership over time. The ideal set of hospitals would be those who have the same executives on their team for the entire sample, as changes could be correlated with the fact that the hospital is penalized. However, this rarely occurs in reality. Therefore, I attempt to proxy for a stable leadership team while still having a large enough sample to run an analysis. Since my focus is clinical experience, my preferred specification is to limit the sample to hospitals who do not change the presence of a doctor on their leadership team from 2010-2014 (2 years before and after the program began). I make this limitation to hospitals in the sample and am left with 524 stable hospitals over time. Finally, I aggregate the information up to the hospital level, where an observation captures whether the hospital has clinical experience on their executive team over the course of the sample.

@@

    I merge the leadership information with the AHA survey using the EIN-AHA links I created. From AHA, I gather the number of beds in the hospital and the Medicare number, which I use to merge the Hospital Compare data. Hospital Compare contains multiple data sets which I use first to limit to hospitals that were penalized in the Hospital Readmissions Reduction Program (HRRP), and also for information on outcome variables: readmission and mortality rates for the relevant HRRP conditions: heart failure, heart attack (AMI), and pneumonia.

    The Hospital Compare excess readmissions files contain information on the excess readmission measure used by CMS to determine whether a hospital should be penalized. Specifically, CMS calculates the average readmission rate of all hospitals, adjusts for demographic factors, and determines the amount of readmissions a hospital has in excess to of the average. This variable is condition specific, where a hospital is penalized if any of the three excess readmission ratios are greater than one\footnote{Additional conditions were added to the penalty calculations in 2015, but I do not consider these since my sample does not extend past 2015}. Further, Hospital Compare also publishes hospital level readmission and mortality rates for heart failure, AMI, and pneumonia. For my main outcomes of interest, I take an average across conditions and weight it using the total number of patients seen for that condition. For further analysis, I create samples of hospitals who were penalized for specific conditions, and investigate the change in readmission and mortality rates for the condition they were penalized in.

@@

    In Table \ref{tab1:sumstats}, I present summary statistics for various hospital characteristics including which condition hospitals are penalized for, and the readmission and mortality outcome variables. After limiting to penalized hospitals with stable leadership teams, my sample consists of between 302-310 hospitals from 2009-2015, except in the case of AMI outcomes, which is only 225-264 hospitals due to small patient numbers. Fourteen percent of hospitals have someone with clinical experience on their executive team, and the size of the hospitals in the sample varies quite a bit. The average number of beds for a hospital in the sample is 186, only slightly lower than the national average for all nonprofits. Hospitals in the sample are less likely to be penalized for AMI than either heart failure or pneumonia. On average, 21\% of patients with one of these conditions would be readmitted. While readmission rates for the specific conditions are similar, the highest likelihood of readmission is for heart failure patients. Alternatively, AMI is the most dangerous condition in terms of mortality rates.

    Figure \ref{fig:weighted_read_mort_graph} shows average readmission and mortality rates across time and across the three different conditions, along with the weighted average. There are multiple notable points the raw data reveals. First, prior to the penalty, hospitals without clinical executives had consistently higher readmission rates than clinical executive hospitals across all conditions. While there is a gap between clinical team and non-clinical team readmission rates, the lines seem to move mostly parallel. In contrast, 2012 begins a decline in readmission rates over all conditions for all types of hospitals. This makes sense and is consistent with previous literature on HRRP, as hospitals have incentive to reduce readmissions in order to reduce future penalties. However, non-clinical hospital teams decrease readmissions at a faster rate, quickly converging with the rates of clinical team hospitals. While suggestive, this motivates further analysis into the differential responses between clinical and non-clinical leadership teams. On the right panel, average mortality rates over time reveal no clear change due to the penalty, and no consistent pattern in the difference between clinical and non-clinical hospital teams.

@@

@@

    \section{Model and Empirical Strategy}

    Not knowing the true form of nonprofit hospital objective functions, we can assume that they maximize some weighted average of revenue and output. Abstracting away from quantity for the purpose of this paper, we can think of firms behaving according to:

    $$\max_{\theta}\hspace{2mm} \delta R + (1-\delta) u(\theta),$$

    \noindent where nonprofits differ from for-profits in that $\delta\neq 1$. Then $R$ is net revenue, $u(.)$ is an increasing function that captures utility from quality of care, and $\theta$ is some quality of care measure. In this setting, we care about hospital responding to a penalty by adjusting readmission rates, so $\theta = 1-r$, where $r$ is the readmission rate. While previous research has set out to understand the average $\delta$ for nonprofit hospitals, $\delta$ may also be dependent on the leadership teams within nonprofit hospitals. Further, introducing the HRRP means that $R$ is directly dependent on $\theta$. We can then write the hospital's problem as

    $$\max_{\theta}\hspace{2mm} \delta(L) R(\theta) + (1-\delta(L)) u(\theta)$$

    \noindent where $L$ captures the composition of the leadership team. Therefore, when maximizing this function, the choice of $\theta$ is dependent on $L$ through its impact on how the hospital values profit vs. quality of care:

    $$\theta = f(\delta(L)).$$

>>>>

    \begin{abstract}
		{\small
        In many industries, including hospitals, nonprofits make up the majority of firms in the market. Understanding the objectives and behaviors of these firms is important when creating incentives. However, nonprofit firms have diverse goals and missions that make understanding their objectives difficult. For hospitals in particular, there is mixed evidence on whether nonprofit hospitals differ meaningfully from for-profit hospitals. In this paper, I investigate whether leadership team composition of nonprofit hospitals affects how similarly they behaves to for-profits. I use a novel data set on nonprofit hospital executives in the US from 2009-2015 to investigate whether the existence of clinical training on a hospital executive team affects hospital behavior. In particular, I leverage several pay-for-performance initiatives in the US in 2012, and estimate the difference in response between for-profits, nonprofits, and nonprofits with clinically trained executives. I find that while readmissions decreased on average for all types of hospitals, the response was larger for for-profits and nonprofits without clinically trained executives. This is consistent with theoretical predictions under the assumption that nonprofits with clinically trained executives put higher weight on societal outcomes than those without clinically trained executives.
		} 
	\end{abstract}

@@

  The objectives and behaviors of firms that are not classic for-profits are not easily determined, and an extensive literature has sought to understand these types of firms. Nonprofits are thought to gain utility not solely from monetary profit, but also through accomplishing societal benefit. Researchers have speculated a variety of objective functions that could motivate nonprofits such as maximizing prestige, income, or a quality/quantity trade-off (\cite{steinberg1986revealed}). Empirical contributions to this literature center around investigating behaviors that differ between for-profit, private nonprofit, and government firms (\cite{sloan2000not}). In this paper, I investigate whether differences in observed behavior between for-profit and nonprofit hospitals vary differentially by nonprofit leadership team composition.

  Hospitals have been a common focus when seeking to understand nonprofit firms. This is because, first, there are different hospital ownership types within the industry, giving researchers the ability to compare the behaviors of nonprofits with that of for-profits. Second, private nonprofit hospitals make up 50\% of all hospitals in the US, and staff, on average, 207 beds. For comparison, for-profits make up around 36\% of hospitals and staff 107 beds (\cite{ASPE_2023}). Thus, nonprofit behavior in the hospital industry is relevant, and directly affects consumers of health care. However, it remains unclear what objective function drives nonprofit hospital behavior, and whether nonprofit hospitals are meaningfully different from for-profits (\cite{sloan2000not}; \cite{erus2002inferring}; \cite{deneffe2002not}; \cite{horwitz2009hospital}).

  While the characterization of being for-profit or nonprofit is certainly important, there could be other characteristics that drive objectives and behavior that are not perfectly correlated with ownership status. Particularly, firm leadership varies widely across firms and potentially drives differences in objectives even within nonprofits. In large, publicly traded firms, executive characteristics are shown to be correlated with firm performance, indicating that there is more to firm behavior than for-profit or nonprofit status (\cite{bertrand2003managing}; \cite{matsa2013female}; \cite{ahern2012changing}). If hospital objectives differ in the weights placed on profit vs. societal benefit, then ownership status certainly plays a role in these weights, but characteristics of those managing the hospital could also play an important role. I contribute to our understanding of hospital and nonprofit behaviors by investigating whether occupational background, whether executives have clinical training, affects response to pay-for-performance incentives.

  Using publicly available tax forms, I construct a novel data set of nonprofit hospital executives from 2009-2015. I link hospitals in this data to characteristics in the American Hospital Association survey and Hospital Compare data. I leverage various pay-for-performance initiatives enacted in 2012 to investigate differential hospital responses between for-profits, nonprofits without clinically trained executives, and nonprofits with clinically trained executives. While executive teams change over time, I find no evidence that they change in response to the programs. This allows me to investigate the role of particular executives in two ways. First, I analyze the differential response to incentives of nonprofits who ever hire a clinically trained executive, as this might be a signal of underlying preferences. Second, I analyze the differential response of nonprofits who have a clinically trained executive in 2012 compared to those who have a clinically trained executive at other times, but not in 2012. These two analyses aim to disentangle the effect of the actual leadership team from the underlying preferences of the hospital.

  I estimate how the interaction between hospital type and post program enactment affects readmission and mortality rates among different groups. Under the assumption that hospitals would have continued to behave similarly in the absence of the programs, and that leadership teams do not change endogenously with the programs, I identify the differential response to pay-for-performance incentives by different hospital types. Among the hospitals in the data, hospitals with clinically trained executives are the minority, and differ on average from other hospitals along various dimensions. Thus, my preferred specification is synthetic difference-in-differences, to analyze a group of comparison hospitals with similar characteristics to treatment hospitals.\footnote{I also present results for the full sample, and with a pre-analysis matching procedure and find qualitatively similar results.}

   While all hospital types decrease readmission rates after the programs are enacted, for-profits and nonprofits without clinically trained executives decrease readmissions more than nonprofits without clinically trained executives. That is, nonprofits without clinically trained executives respond to incentives more similarly to for-profits than other nonprofits. To investigate the mechanisms by which hospitals decrease readmissions, I also conduct the analyses with uncompensated care and case mix index as outcomes. (results of this analysis here)

    This paper contributes to three strands of literature. First, it adds to our understanding of what underlies nonprofit hospital behavior. This literature is extensive, and a review of it is beyond the scope of this paper. To summarize, there are various aspects of nonprofit behavior and objectives for which the evidence of how much it differs from for-profits remains mixed: costs, uncompensated care, technology adoption, and quality (\cite{sloan2000not}; \cite{eggleston2008hospital}; \cite{moscelli2018effect}; \cite{moscone2020public}). What underlies this is the need to understand the ways in which nonprofits make choices differently than for-profits. To this point, researchers have found that nonprofits do not act as purely profit maximizers, but maximize some combination of profit and output (\cite{deneffe2002not}, \cite{chang2011nonprofit}). A paper with an approach similar to mine which investigates hospital behavior is \citeauthor{chang2011nonprofit} (\citeyear{chang2011nonprofit}), which looks at the effect of a cost shock on likelihood of shutting down and mix of profitable services. They find that nonprofits and for-profits are equally as likely to shut down after the shock, but only nonprofits adjust their mix of profitable services. Similarly, I investigate a change in incentives to hospitals, but I look at the differential impacts of leadership teams within nonprofits. This paper adds to our understanding of inputs into nonprofit hospital objective functions.

    Second, many researchers have documented a correlation between executive backgrounds, specifically for CEOs, and firm performance, starting with \citeauthor{bertrand2003managing} (\citeyear{bertrand2003managing}), who show that manager fixed effects are an important driver of many firm behaviors and decisions. Female board members are associated with better oversight and more women executives (\cite{matsa2011chipping}; \cite{adams2009women}), but are negatively correlated with firm value (\cite{ahern2012changing}). Having more female executives is correlated with choices of female employee wages and corporate strategies (\cite{flabbi2019female}; \cite{matsa2013female}); young male CEOs tend to be more aggressive in mergers and acquisitions, while those with military experience are less aggressive (\cite{levi2010deal}; \cite{benmelech2015military}); CEOs with general ability tend to receive higher pay and perform better (\cite{kaplan2012ceo}; \cite{custodio2013generalists}, \cite{adams2018director}; \cite{frydman2019rising}). Further, Chief Diversity Officers are not found to have any effect on hiring more diversely in universities (\cite{bradley2022impact}). Due to data limitations, many of these studies focus on large, publicly traded firms in the US. However, one study uses data from England to study whether CEOs affect hospital production, and find no association (\cite{janke2019impact}). \textcolor{red}{(add paper about Chile hospital executives)}. I contribute to this literature first by investigating a context which is not largely studied but is policy relevant, nonprofit US hospitals, and considering a unique executive characteristic pertinent to that context.

    Finally, I contribute to our understanding of how providers respond to pay-for-performance incentives in health care. The most in depth study of how hospitals respond to HRRP is \citeauthor{gupta2021impacts} (\citeyear{gupta2021impacts}). The author finds that hospitals decreased readmissions by 5\% and mortality rates by 2\% on average as a result of the program, confirming prior studies (\cite{mellor2017does}; \cite{ziedan2018essays}; \cite{ody2019decreases}; \cite{gupta2021impacts}). Importantly, around 40\% of the decrease in readmissions is due to selective patient practices. This raises important questions about how to incentivize improvement in quality of care without also incentivizing gamesmanship. This paper addresses a necessary first step in considering whether leadership compositions could affect likelihood of gaming. \textcolor{red}{(add papers about other program too)}

@@

    The Hospital Readmissions Reduction Program (HRRP) was passed as a part of the Affordable Care Act in 2010 as a pay-for-performance incentive for hospitals to increase quality. In October 2011, the Center for Medicare and Medicaid Services (CMS) released a set of rules that mandated penalties for hospitals with above average readmission rates. A readmission is when a patient returns to the hospital within 30 days of being discharged from a previous stay; avoidable readmissions are a bad outcome for patients and increase health care spending. The goal of HRRP is to lower readmissions through better care coordination, less initial stay complications, and better post-care instructions. Beginning in October 2012, hospitals with higher readmission rates than the national average in pneumonia, heart failure, or AMI (after adjusting for demographic characteristics) receive a lower reimbursement rate for Medicare patients. In 2015, CMS also included chronic obstructive pulmonary disease, coronary artery bypass graft surgery, and elective primary total hip arthroplasty and/or total knee arthroplasty as conditions which go into the penalty calculation. \textcolor{red}{(add citations??)}

    Penalized hospitals pay penalties in the form of a fixed rate for every Medicare patient regardless of the condition. Further, CMS does not distinguish a necessary readmission from an avoidable readmission; any repeat hospital visit is included in the penalty calculation. Excess readmission rates are calculated using a rolling look-back period of 3 years to determine whether the hospital is penalized. Therefore, hospitals had incentive to react immediately once details of the program were announced in October of 2011. These penalties are not insignificant; penalized hospitals paid, on average, 4-5\% of revenue.

    \textcolor{red}{(is the program still going?)}

    \subsection{Hospital Value-Based Purchasing Program}

@@

    Nonprofits are governed by a board of directors, whose role is to set high level goals and strategies, and provide general oversight. The board is not in charge of day-to-day operations, but the general direction of the firm. They select members of the executive team to carry out day-to-day operations. While hospital executives typically do specialize by getting degree(s) in healthcare management or an MBA specific to healthcare, they do not often have clinical training. While some doctors earn additional degrees before stepping into an executive role, this is not a necessary condition to becoming a physician executive. An executive team usually consists of at least a CEO and CFO, but every nonprofit is different in exactly how they structure their executive team. Some nonprofits hire Chief Medical Officers, Chief Quality Officers, executive directors, vice presidents of various departments, etc. I consider all of these employees as part of the day-to-day management of the hospital and include them as a part of the executive team.

    All tax-exempt organizations in the US are required to file a Form 990 with Internal Revenue Services (IRS) each year. There are different types of forms, but any organization grossing over \$200,000 must file the most extensive Form 990. Sections of this form include a statement of revenue, statement of functional expenses, a balance sheet, and, as used in this project, a list of all key employees, executives, and board members. In this section, each firm is required to report the name and title, average hours per week, position, and compensation of their board members and executive leaders. Since these firms are tax exempt, their information in these forms is made publicly available.

@@

    To characterize hospitals as having clinically trained executives or not, I construct a novel data set of nonprofit hospitals in the US that contains names and select characteristics of executives tied to the hospital from 2009-2015. I gather this data from each hospital's publicly available Tax Form 990, which every sufficiently large nonprofit files each year in the US. To my knowledge, this is the first large scale gathering of leadership names from these forms.

    The historical tax forms for all nonprofits are housed by ProPublica.\footnote{https://projects.propublica.org/nonprofits/} I use the NonProfit Explorer API to extract the Employee Identification Numbers (EIN) of all nonprofits classified as a hospital according to their National Taxonomy of Exempt Entities (NTEE) code. After filtering out associations and specialty hospitals, there are around 3,000 firms left. Importantly, I need to be able to link hospitals in the tax forms to other hospital characteristics from publicly availble sources. No official crosswalk between EINs and AHA IDs or Medicare numbers exists, so I must rely on matching by location and name of hospital in both data sets. After limiting to only general, nonprofit hospitals that I can link to the AHA survey, I confidently match approximately 1200 EINs to an AHA identification number based on exact name matches within the same state.

    For each of the 1200 EINs, I then extract web URLs to Form 990 PDFs in each year. I download these locally and use optical character recognition (OCR) text extraction methods to scrape the relevant section of the pdf, which is titled ``Officers, Directors, Trustees, Key Employees, and Highest Compensated Employees". Finally, I use string cleaning methods to identify names and positions for each EIN, year. I also identify titles indicating a physician: MD, Dr., or DO. Using OCR extraction can be unreliable in some instances when the text is handwritten or the font is too small. Therefore, I manually find names, positions, and titles for observations that are missing after the initial extraction. This yields 850 EINs that are matched to AHA data, and are not missing text information.

    Due to the lack of consistency in organization of executive teams, I focus mainly on the executive teams as a whole instead of specific positions. I drop anyone who is solely a board member, and focus on those who have executive in their title or are labeled as a president or vice president of a specific aspect of the hospital. I create various hospital-level characterizations based on their executive teams, such as the existence of a clinically trained executive, the number of clinically trained executives, the number of total executives, and whether the hospital has a Chief Medical Officer.

@@

    I include hospital information from the AHA survey on the number of beds in the hospital and the Medicare number, which I use to merge the Hospital Compare and Healthcare Cost Reporting Information System (HCRIS) data. Further, I use the AHA categorization of for-profit or nonprofit. Hospital Compare contains information on penalties from HRRP and readmission and mortality rates for the relevant HRRP conditions: heart failure, heart attack (AMI), and pneumonia. From HCRIS, I gather information on penalties from the HVBP program, operating expenses, and uncompensated care. I also link to the publicly available CMS file containing each hospital's case mix index.

    The Hospital Compare excess readmissions files contain information on the excess readmission measure used by CMS to determine whether a hospital should be penalized. Specifically, CMS calculates the average readmission rate of all hospitals, adjusts for demographic factors, and determines the amount of readmissions a hospital has in excess to of the average. This variable is condition specific, where a hospital is penalized if any of the three excess readmission ratios are greater than one\footnote{Additional conditions were added to the penalty calculations in 2015, but I do not consider these since my sample does not extend past 2015}. Further, Hospital Compare also publishes hospital level readmission and mortality rates for heart failure, AMI, and pneumonia. For my main outcomes of interest, I take an average across conditions, weighted by the total number of patients seen for that condition.

@@

    In Table \ref{sumstats}, I present summary statistics for hospital variables. There are 3,766 total hospitals in the sample, with 150 beds on average. Hospitals in the sample are less likely to be penalized for AMI than either heart failure or pneumonia under the HRRP program. On average, 20\% of patients with one of these conditions would be readmitted. While readmission rates for the specific conditions are similar, the highest likelihood of readmission is for heart failure patients. Alternatively, AMI is the most dangerous condition in terms of mortality rates.

    %\import{Tables}{overall_sumstats.tex}

    I present a table of means for each hospital type (for-profit, all nonprofits, nonprofits with a clinically trained executive, and nonprofits without a clinically trained executive) in Table \ref{tab:sumstats_samples}. Eighty-five percent of hospital in the sample are classified as nonprofits. Importantly, the two sub-samples of nonprofits do not add up to the total number of nonprofits because of the limited ability to extract leadership text for all nonprofits. There are key differences in nonprofits with and without clinically trained executives: those with MDs are larger, more likely to be penalized, have higher uncompensated care, and more complex patients.

    %\import{Tables}{sample_sumstats.tex}

    I plot readmission and mortality rates across time for different hospital types in Figure \ref{fig:weighted_read_mort_graph}. While this is purely descriptive evidence, the trends in outcomes point to more similar behavior between for-profits and nonprofits without clinically trained executives in comparison to nonprofits with a clinically trained executive.

@@

    Finally, I present a table of correlations between having at least one clinically trained executive and other time invariant hospital characteristics. There may be a concern that this particular characteristic of executive teams is simply a proxy for another quality, such as being an academic medical center. Unsurprisingly, the most correlated characteristic with having a clinically trained executive is having a CMO on the executive team. However, even this correlation is only .4, meaning there is still variation in having clinically trained executives even when controlling for having a CMO.

@@

    \section{Model and Empirical Strategy}\label{sec:model}

    \subsection{First Period}

    Not knowing the true form of hospital objective functions, we can consider them maximizing some weighted average of profit and societal (or patient) benefit. Abstracting away from quantity and price as functions of quality for the purpose of this paper, I present a two-stage model of firm behavior that captures the change of quality due to pay-for-performance incentives. In the first period, prior to program enactment, hospitals do not expect any penalties in the future, and choose quality according to:

    $$\max_{\theta}\hspace{2mm}\alpha\left[R - c_{\theta}(\theta) \right] + (1-\alpha) u(\theta),$$

    \noindent where $R$ is net revenue, $c_{\theta}(.)$ is an increasing cost function specific to quality $\theta$, $u(.)$ is extra utility, which is increasing and concave is $\theta$, and $\alpha\in[0,1]$ is a hospital's type, indicating how much they care about profit vs. societal benefit. There is an implicit stay open condition throughout the model. Taking the first order condition yields

    $$(1-\alpha)u'(\theta) = \alpha c_{\theta}'(\theta),$$

    \noindent marginal benefit of increasing quality equals marginal cost. Solving for $u'(\theta)$ and differentiating with respect to $\alpha$ yields

    $$\frac{du'(\theta)}{d\alpha} = c_{\theta)}'(\theta)\frac{1}{(1-\alpha)^2} > 0.$$

    Thus, by the Implicit Function Theorem,

    $$\frac{d\theta}{d\alpha} = \frac{du'(\theta)/d\alpha}{du'(\theta)/d\theta} < 0.$$

    That is, the more weight the hospital places on profit, the lower quality of care they will provide.

    \subsection{Second Period}

    Hospitals determine period 1 quality $\theta_1$ according to $\theta(\alpha)$ described above. Then, in period two, hospitals are penalized according to $\theta_1$, and they choose a new quality, $\theta_2$, which will determine penalties in the future. The penalty $(k)$ comes in the form of a negative shock to revenue for hospitals with quality below a certain threshold. That is,

    $$R_2 = \begin{cases}
        R - k, \text{  if  } \theta_1<\overline{\theta}\\
        R, \text{  if  } \theta_1\geq \overline{\theta}.
    \end{cases}$$
